---
title: "Problem Session 2"
subtitle: "Loops, vectorization, and functions"
format: html
editor: visual
---

## For loops with lists

Let's start with a `for` loop that does something a bit more interesting than printing values. We are going to combine three things from the week's material: a `for` loop, subsetting a data frame, and building a named list as we go.

Recall that the `iris` data set has 150 rows across three species (`setosa`, `versicolor`, `virginica`).

```{r}
head(iris)
levels(iris$Species)
```

Our goal:

-- Loop over each species in `iris`

-- For each species, compute the **mean** and **standard deviation** of `Petal.Length`, plus the **number of observations**

-- Store these in a named list where each element is itself a list of three summary values

Before writing the loop, take a moment to think about how you would initialize an empty list, and how to add to it as the loop runs. We've seen that `[[ ]]` is the way to grab (or assign to) a single element of a list by name.

```{r}
species_summary <- list() #initializing an empty list

for (sp in unique(iris$Species)) {
  subset_data <- iris[iris$Species == sp, ]
  species_summary[[sp]] <- list(
    mean_pl = mean(subset_data$Petal.Length, na.rm = T),
    sd_pl = sd(subset_data$Petal.Length, na.rm = T),
    n = nrow(subset_data)
  )
}

species_summary

species_summary$versicolor
species_summary[["versicolor"]]
```

A few things worth discussing once we have it working:

-- What does `unique(iris$Species)` return, and why is looping over that cleaner than looping `1:3`?

-- How would you pull out just the mean for `versicolor` from `species_summary`?

## Vectorization

The previous problem felt manual. R gives us vectorized functions that run the loop under the hood in compiled C code, which is faster and easier to read.

For this problem, our goal is to compute the **column means** for the four numeric columns of `iris` and compare:

-- A `for` loop version (pre-allocate a numeric vector, fill it in)

-- A vectorized version using `colMeans()`

Then we will compare the computing time between the two approaches. Note that we use the `microbenchmark` package, which you may need to install with `install.packages("microbenchmark")`; if you haven't already done so.

```{r}
library(microbenchmark)

numeric_cols <- iris[, 1:4] # creating a data.frame with numeric columns only

# Loop version
col_means_loop <- numeric(ncol(numeric_cols)) # initialized an empty vector

microbenchmark(for (i in seq_along(numeric_cols)){
  col_means_loop[i] = mean(numeric_cols[[i]])
})


# Vectorized version
colMeans(numeric_cols)

names(col_means_loop) <- names(numeric_cols)
col_means_loop
microbenchmark(colMeans(numeric_cols))
```

```{r}
numeric_cols
numeric_cols[3]
numeric_cols[[3]]
```

## Vectorized if/then/else

The same idea applies to conditional logic. We have seen `if/else if/else` inside a `for` loop, and we have seen `ifelse()` which is vectorized. Let's practice using both.

Create a new column `size` in `iris` based on `Sepal.Length`:

-- `"small"` if length $\lt$ 5

-- `"medium"` if 5 $\leq$ length $\leq$ 6.5

-- `"large"` if length $\gt$ 6.5

Do it two ways: first with a `for` loop and `if/else if/else`, then with nested `ifelse()`. Confirm both give the same result.

```{r}
# for loop with if/ else if/else
head(iris)

iris$size_loop <- character(nrow(iris)) # initialized

for (i in seq_len(nrow(iris))) {
  if (iris$Sepal.Length[i] < 5) {
    iris$size_loop[i] <- "small"
  } else if (iris$Sepal.Length[i] <= 6.5) {
    iris$size_loop[i] <- "medium"
  } else {
    iris$size_loop[i] <- "large"
  }
}

head(iris)

# Vectorized with ifelse
iris$size_vectorized <- ifelse(iris$Sepal.Length < 5, "small", ifelse(iris$Sepal.Length <= 6.5, "medium", "large"))

head(iris)
```

## Writing a function with `stop()` and `switch()`

Now let's write our own function. The idea: write a function called `make_table()` that builds a frequency table from a categorical column of a data frame. We want it to:

-- Take a data frame and a column name (as a string) as inputs

-- Have a `type` argument that controls the output: a plain table, a proportions table, or a sorted (descending) table

-- Use `stop()` to validate inputs: the first argument must be a data frame, and the named column must exist

-- Use `switch()` to dispatch on `type` rather than a chain of `if/else` statements

-- Drop `NA` values before tabulating

Default `type` to `"table"` so the user doesn't have to specify it for the common case.

```{r}
make_table <- function(df, col, type = "table"){
  if(!is.data.frame(df)) {
    stop("`df` must be a data frame.")
  }
  if(!(col %in% names(df)) ) {
    stop("Column ", col, " not found in data frame")
  }
  
  x <- df[[col]]
  x <- na.omit(x) #remove missing values
  
  switch(type,
         table =table(x),
         prop = prop.table(table(x)),
         sorted = sort(table(x), decreasing = T),
         stop("`type` must be one of: 'table', 'prop', 'sorted'.")
  )
}

# try it out
make_table(df = iris, col = "Species")
make_table(df = iris, col = "size_loop", type = "prop")
make_table(iris, "size_vectorized", "sorted")

#make_table(1:10, "Species")
#make_table(iris, "Species", "prop")
```
