---
title: "Problem Session 2"
subtitle: "Loops, vectorization, and functions"
format: html
editor: visual
---

## For loops with lists

Let's start with a `for` loop that does something a bit more interesting than printing values. We are going to combine three things from the week's material: a `for` loop, subsetting a data frame, and building a named list as we go.

Recall that the `iris` data set has 150 rows across three species (`setosa`, `versicolor`, `virginica`).

```{r}

```

Our goal:

-- Loop over each species in `iris`

-- For each species, compute the **mean** and **standard deviation** of `Petal.Length`, plus the **number of observations**

-- Store these in a named list where each element is itself a list of three summary values

Before writing the loop, take a moment to think about how you would initialize an empty list, and how to add to it as the loop runs. We've seen that `[[ ]]` is the way to grab (or assign to) a single element of a list by name.

```{r}

```

A few things worth discussing once we have it working:

-- What does `unique(iris$Species)` return, and why is looping over that cleaner than looping `1:3`?

-- How would you pull out just the mean for `versicolor` from `species_summary`?

## Vectorization

The previous problem felt manual. R gives us vectorized functions that run the loop under the hood in compiled C code, which is faster and easier to read.

For this problem, our goal is to compute the **column means** for the four numeric columns of `iris` and compare:

-- A `for` loop version (pre-allocate a numeric vector, fill it in)

-- A vectorized version using `colMeans()`

Then we will compare the computing time between the two approaches. Note that we use the `microbenchmark` package, which you may need to install with `install.packages("microbenchmark")`; if you haven't already done so.

```{r}

```

## Vectorized if/then/else

The same idea applies to conditional logic. We have seen `if/else if/else` inside a `for` loop, and we have seen `ifelse()` which is vectorized. Let's practice using both.

Create a new column `size` in `iris` based on `Sepal.Length`:

-- `"small"` if length \< 5

-- `"medium"` if 5 ≤ length ≤ 6.5

-- `"large"` if length \> 6.5

Do it two ways: first with a `for` loop and `if/else if/else`, then with nested `ifelse()`. Confirm both give the same result.

```{r}

```

## Writing a function with `stop()` and `switch()`

Now let's write our own function. The idea: write a function called `make_table()` that builds a frequency table from a categorical column of a data frame. We want it to:

-- Take a data frame and a column name (as a string) as inputs

-- Have a `type` argument that controls the output: a plain table, a proportions table, or a sorted (descending) table

-- Use `stop()` to validate inputs: the first argument must be a data frame, and the named column must exist

-- Use `switch()` to dispatch on `type` rather than a chain of `if/else` statements

-- Drop `NA` values before tabulating

Default `type` to `"table"` so the user doesn't have to specify it for the common case.

```{r}

```
