---
title: "Problem Session 3"
subtitle: "Reading Data, dplyr, tidyr & SQL"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Setup

We'll use the `tidyverse` and `nycflights13` packages throughout this session. Make sure they're installed before running the code below.

```{r}
library(tidyverse)
library(nycflights13)
```

# Part 1: Reading Data

## 1.1 Reading a Delimited File with `read_delim()`

We'll start by reading in the `household_power_consumption.txt` dataset from the UC Irvine Machine Learning Repository <https://archive.ics.uci.edu/dataset/235/individual+household+electric+power+consumption>. In this case we can use the menus to read in the data we saved locally. If you open the raw file, you'll notice the values are separated by semicolons (`;`) rather than commas — this is a **semicolon-delimited** file.

If we go to File -\> Import Dataset we can choose to import a "From Text" data using `base` R or the `readr` method. We chose the `readr` way and checked out the default method. We had to change our delimiter to a ";" and the data should read in.

The code from the import menu is produced here.

```{r}

```

What do you notice? Many numeric columns are read in as character (`<chr>`). Let's investigate why.

```{r}

```

The `?` character is being used to represent missing values in this dataset, which confuses `readr`. We can fix this using the `na =` argument:

```{r}

```

Always inspect your data after reading it in. Unexpected character types are often a sign that missing values aren't being handled correctly.

## 1.2 Tibble vs. Data Frame

Before we dive into manipulation, let's quickly note that `read_csv()` and `read_delim()` return **tibbles**, not plain data frames. Tibbles are a tidyverse-friendly version of a data frame with some helpful differences:

-   They print only the first 10 rows and as many columns as fit on screen.
-   They never silently convert strings to factors.
-   They never create row names.

Let's look at the differences in the output when we type `flights` (tibble) in the console versus typing `mtcars` (data frame) in the console.

You can coerce a regular data frame to a tibble with `as_tibble()`. Let's use `as_tibble()` on the `mtcars` data set. Save this new data set as `mtcars_tibble`. Check your environment to see if it worked!

```{r}

```

# Part 2: Data Manipulation with `dplyr`

**dplyr** is the primary package in the tidyverse for data wrangling. We already loaded tidyverse package and it took care of dplyr package as well.

For the rest of this session we'll use `flights` from the `nycflights13` package — records of every flight departing New York City in 2013.

```{r}
glimpse(flights)
```

## 2.1 `select()` — Choosing Columns

### Question 1 (Warm-up)

Select only the `carrier`, `origin`, `dest`, and `air_time` columns from `flights`.

```{r}

```

### Question 2

Make a data frame that keeps every variable except `dep_delay`. Call the new data frame `new.data`.

```{r}

```

### Question 3

Use `select()` with a range (`:`) to keep all columns from `year` through `arr_delay` (inclusive). These are the scheduling and delay variables.

```{r}

```

### Question 4

Use `select()` with the `contains()` helper to keep only columns whose name includes the string `"dep_"`. What columns do you get?

```{r}

```

## 2.2 `filter()` — Choosing Rows

### Question 5

Filter `flights` to only include flights that departed from JFK airport.

```{r}

```

### Question 6

What if we want to look at destinations of RDU and JFK that had a departure delay greater than 60 minutes? Write the code below to do this...

```{r}

```

Why `c`?

What happens if we change %in% to == ?

## 2.3 `arrange()` — Sorting Rows

### Question 7

Arrange the `flights` data by `dep_delay` in descending order so the most delayed flights appear at the top. What is the longest departure delay, and what carrier operated that flight?

```{r}

```

## 2.4 Chaining: `filter()` + `arrange()` + `select()`

### Question 8

Create a data frame that:

1.  Keeps only flights to RDU.
2.  Arranges them by departure delay (largest first).
3.  Selects only `tailnum`, `carrier`, `dep_delay`, and `arr_delay`.

What is the tail number of the most delayed flight to RDU?

```{r}

```

## 2.5 `mutate()` — Creating New Variables

### Question 9

Create a new variable `speed_mph` that gives the average speed of each flight in miles per hour. Recall that `air_time` is in minutes and `distance` is in miles. Keep only `flight`, `carrier`, `distance`, `air_time`, and `speed_mph`.

```{r}

```

### Question 10

Use `mutate()` and `ifelse()` to create a new column `on_time` that is `"On Time"` if `arr_delay <= 0` and `"Late"` otherwise. Then use `count()` to find how many flights fall into each category.

```{r}


```

## 2.6 `group_by()` + `summarise()` — Summary Statistics by Group

### Question 11

Find the **mean and standard deviation of departure delay** for each carrier. Arrange the result by mean departure delay in descending order. Which carrier has the worst average departure delay?

```{r}

```

### Question 12 (Challenge)

For each month, compute:

-   The total number of flights.
-   The mean arrival delay.
-   The proportion of flights that arrived on time.

Which month has the highest proportion of on-time arrivals?

```{r}


```

# Part 3: Reshaping Data with `tidyr`

For this section we'll build a small summary tibble from `flights` and then practice reshaping it.

```{r}
# Build a mean arrival delay table by carrier and month (first 3 months)
delay_summary <- flights |>
  filter(month %in% 1:3) |>
  group_by(carrier, month) |>
  summarise(mean_arr_delay = round(mean(arr_delay, na.rm = TRUE), 1),
            .groups = "drop")

delay_summary
```

## 3.1 `pivot_wider()` — Long to Wide

### Question 13

Pivot `delay_summary` so that each row is a **carrier** and there is one column per month (`month_1`, `month_2`, `month_3`). Save this as `delay_wide`.

```{r}

```

## 3.2 `pivot_longer()` — Wide to Long

### Question 14

Take `delay_wide` and pivot it back to long format so that we have columns `carrier`, `month`, and `mean_arr_delay`. The result should match the original `delay_summary`.

```{r}

```

## 3.3 Putting It All Together

### Question 15 (Challenge)

Starting from `flights`, create a **wide** table showing the **count of flights** for each combination of `origin` airport (rows) and `carrier` (columns). The result should have one row per origin airport and one column per carrier.

```{r}

```

# Google's Big Query

Lastly, let's connect to Big Query. This is an cloud resource from google. They have some example large datasets up there and you can also put your data up there if you want.

We have to start by setting up a project. Once set up (no credit card entered) you get a project ID that you can use to do some querying!

## Direct SQL Query via `bigrquery`

This code is modified from <https://cloud.google.com/vertex-ai/docs/workbench/user-managed/use-r-bigquery>. Once each R session, you need to run this code in the console so you can log in (unless you take care of authentication in another way). You'd need to install the bigrquery package if you haven't done so.

```{r}

```

## `dplyr` Interface via `DBI`

This code is modified from <https://bigrquery.r-dbi.org/> where they describe the `dplyr` interface for dealing with BigQuery data tables.

```{r}

```
