---
title: "Problem Session 4"
subtitle: "Exploratory Data Analysis"
format: html
editor: visual
---

# Packages

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
library(nycflights13)
```

We'll work with two datasets today:

-   `penguins` from `palmerpenguins` — measurements on 344 penguins across three species
-   `flights` from `nycflights13` — over 300,000 flights departing NYC in 2013

# Contingency Table

We'll use the `flights` dataset to practice creating contingency tables. First, use `glimpse()` to remind yourself what variables are available.

```{r}

```

## One-way table

Create a one-way contingency table showing the number of flights from each `origin` airport. Try both the Base R `table()` approach and the tidyverse approach.

```{r}
# Base R


# tidyverse


```

## Two-way table

Now cross `origin` with `carrier`. Use `group_by()` + `summarise()` and then `pivot_wider()` to display it in wide format with carriers across the top.

```{r}

```

# Numerical Summaries

Let's use the `penguins` dataset to practice numerical summaries.

Compute the mean and standard deviation of `body_mass_g` for each species. Don't forget to handle missing values!

```{r}

```

Now extend this using `across()` to compute the same statistics for all numeric columns at once, still grouped by species.

```{r}

```

# ggplot

Now let's explore the `penguins` dataset with some plots.

## Histogram

Make a histogram of penguin body mass. Set an appropriate binwidth. Pull up the help file for `geom_histogram()` if you need a reminder on the argument name.

```{r}

```

## Scatterplot

Create a scatterplot of `body_mass_g` (x) vs. `flipper_length_mm` (y).

```{r}

```

Note: aesthetic is a visual property of one of the objects in your plot. Aesthetic options are:

-- shape

-- color

-- size

-- fill

**Question**: What happens if we put `color` as an argument directly in our `geom` instead of inside `aes()`? Try setting `color = "blue"` inside `geom_point()`, then try mapping `color = species` inside `aes()`. What's the difference?

## Formatting

What if your legend goes off the page (this may be an issue on your project...)? We can use the `guides()` function to control it. Try adding the following to your scatterplot:

`guides(color = guide_legend(ncol = 2))`

Talk through what each part of this code is doing, then add it to your scatterplot above.

For an additional resource, see [here](https://www.tidyverse.org/blog/2024/02/ggplot2-3-5-0-legends/).

## Labels

Good labels are critical for making your plots accessible to a wider audience. Always ensure the axis and legend labels display the full variable name.

`labs()` is the function. Common arguments to set are:

-- x

-- y

-- title

-- subtitle

-- caption

```{r}

```

What if your axis labels are long? Sometimes it's better to rotate them! We can do that using the `theme()` function:

`theme(axis.text.x = element_text(angle = 90, vjust = 0.5))`

```{r}

```

What if your title is too long? We can fix this by putting `\n` in your title text. Try it!

```{r}

```

## Custom Themes

We can use the `theme()` function for a lot of things — font, color, bold text, and more. One powerful trick: we can save a theme as an R object and add it to any plot with `+`.

```{r}
wolftheme <- theme(
  plot.title   = element_text(family = "Helvetica", face = "bold", size = 20),
  legend.title = element_text(colour = "red", face = "bold.italic", family = "Helvetica"),
  legend.text  = element_text(face = "italic", colour = "darkred", family = "Helvetica"),
  axis.title   = element_text(family = "Helvetica", size = 10, colour = "darkred"),
  axis.text    = element_text(family = "Courier", colour = "black", size = 10)
)

# save your scatterplot as an R object and add the wolf theme to it using +
```

# Break Apart a Plot by a Variable

Consider this histogram below:

```{r}
penguins |>
  ggplot(aes(x = body_mass_g, fill = species)) +
  geom_histogram(binwidth = 200, alpha = 0.5)
```

What if we don't want the overlap? We can use `facet_wrap()` to split the plot into panels! This function takes the name of the variable you want to split by, and how many cols/rows you want. Note: the syntax for this function is `~variable.name`.

Run `?facet_wrap` in your console to see the name of the row and column arguments.

Add `facet_wrap()` as a layer below, breaking the plot apart by species.

```{r}

```

## Break Apart by More Than One Variable

Now let's make a scatterplot between `bill_length_mm` and `bill_depth_mm`. Color the points by species and shape the points by island.

```{r}

```

Here, we are going to introduce `facet_grid()`. It's a lot like `facet_wrap()`, but allows us to break our plots apart by more than one variable. The syntax for `facet_grid()` is `variable1 ~ variable2`. Try breaking the plot above by both species AND island.

```{r}

```

For more information on the difference, please reference this [stackoverflow link](https://stackoverflow.com/questions/20457905/whats-the-difference-between-facet-wrap-and-facet-grid-in-ggplot2).

## Fill vs. Color

Change `fill` to `color` below and re-run the code. What happened?

```{r}
penguins |>
  ggplot(aes(x = body_mass_g, fill = species)) +
  geom_histogram(binwidth = 200, alpha = 0.7) +
  facet_wrap(~species, nrow = 3)
```

Generally, `fill` defines the color *inside* a geom, whereas `color` defines the color of the *outline*. However, there is one exception to the rule... points!

Most points plotted are considered too small to have meaningful area inside, so `fill` won't work! Try it below:

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point()
```

`fill` and `color` do act differently when points have a "border". Shapes 21–25 are shapes that have a border. Use the `shape` argument within `geom_point()` to change the shape, and see how it interacts with `fill` and `color`.

# Recreating a Plot

Let's start by reading in some data from a URL. This is a `.csv` file so it is a comma delimited file. Opening it, we saw that there is a header row and some missing values. But it seems like it should read in just fine with `read_csv()`.

```{r}

```

The file at <https://www4.stat.ncsu.edu/~online/datasets/sleep_data_info.txt> gives the reference for where this data comes from. Let's recreate the first plot (almost) that we see in that paper.

```{r}

```
