---
title: "Problem Session 5"
subtitle: "apply, purrr, and Advanced Function Writing"
format: html
editor: visual
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(nycflights13)
```

## The `apply()` Family

### `apply()` and Anonymous Functions with `penguins`

We'll start by practicing `apply()` on the `penguins` data set from the `palmerpenguins` package. Our goal is to compute a custom summary for each numeric variable: we want the mean, standard deviation, and the range (max - min).

```{r}

```

### `lapply()` and `sapply()` with a List of Data Frames

Below we create a list containing two subsets of the `flights` data: one for carrier `"UA"` (United Airlines) and one for `"DL"` (Delta). Both subsets contain only the numeric columns relevant to departure and arrival performance.

```{r}
flights_list <- list(
  UA = flights |> filter(carrier == "UA") |>
         select(dep_delay, arr_delay, air_time, distance),
  DL = flights |> filter(carrier == "DL") |>
         select(dep_delay, arr_delay, air_time, distance)
)
```

**Part A:** Use `lapply()` to compute the column means for each data frame in `flights_list`. What is the structure of the output?

```{r}

```

**Part B:** Now use `sapply()` to do the same thing. How does the output differ from `lapply()`?

```{r}

```

**Part C:** Using `sapply()`, write a single-line call that returns, for each carrier, the proportion of flights with a departure delay greater than 15 minutes. Use an anonymous function inside `sapply()`.

```{r}

```


### `tapply()`: Grouped Summaries

`tapply()` is a member of the apply family that often flies under the radar. It lets you apply a function to subsets of a vector, where the subsets are defined by a grouping factor — essentially a base R version of `group_by() |> summarize()`.

The syntax is:

```         
tapply(X, INDEX, FUN, ...)
```

where `X` is the vector of values, `INDEX` is the grouping factor (or list of factors), and `FUN` is the function to apply.

**Your turn:** Using `tapply()` on the `penguins` data, compute the mean body mass (in grams) for each combination of `species` and `island`. 

```{r}

```

After you have the result, think about what type of object `tapply()` returns when two grouping factors are provided. How does this compare to a `group_by() |> summarize()` approach?

## `purrr`

### `map()` and `map_dbl()` on a List

Let's practice with `purrr`'s `map()` family, which gives us a more consistent and expressive alternative to `lapply()` and `sapply()`.

We'll use the same `flights_list` from above.

**Part A:** Use `map()` to compute the column means for each carrier data frame. Then use `map_dbl()` to extract just the mean `arr_delay` for each carrier in a single step.

```{r}

```

**Part B:** Use `map2()` to compute, for each carrier, the difference between mean arr_delay and mean dep_delay. In other words, on average, do flights make up time in the air, or do delays compound?

*Hint:* You'll need to supply two lists: one containing the mean `arr_delay` for each carrier and one containing the mean `dep_delay`. Or you can work directly with `flights_list` and a two-argument anonymous function.

```{r}

```

### `map_dfr()` and Combining Results

Now let's introduce a particularly useful function from `purrr`: `map_dfr()`. One nice feature of `purrr` over base R's apply family is that we have direct control over the output format. The `dfr` in `map_dfr()` stands for "data frame, row-bound" — it applies a function to each element of a list and automatically stacks the results into a single tibble by binding rows.

Because we're working with data frames in our list, it may (or may not) be logical that we want to combine everything into one working data frame. We'll use the `flights_list` from above.

For this exercise, work through the following three steps:

-- Use `map_dfr()` to calculate the column means for all numeric variables in each data frame in `flights_list`.

-- Create a new variable called `group` that identifies which list element each row came from — values should be `1` for UA and `2` for DL.

-- Pivot the resulting data set into a longer form that would be better suited for plotting or modeling (one row per carrier–variable combination).

```{r}

```

## Advanced Function Writing

### Writing a Pipeable `tidyverse`-Style Function

Write a function called `scale_numeric_cols()` that takes a data frame `.df` and returns the same data frame with all numeric columns standardized (mean 0, SD 1). The function should use `mutate()` and `across()` internally.

Requirements:

-   The function should work in a pipeline.
-   Use `stop()` to throw an informative error if the input is not a data frame.

```{r}

```

Test it:

```{r}

```

### Read Through a Function: `enquo()`, `!!`, and `as_label()`

Below is a function that computes a grouped numeric summary and names the output column dynamically based on the column name the user passes. Read through it carefully and add comments explaining what each non-obvious line does.

Pay special attention to:

-   Why we need `enquo()` and `!!`
-   What `as_label()` gives us
-   The walrus operator `:=`

```{r}
library(tidyverse)

grouped_summary <- function(.df, group_var, numeric_var, na.rm = TRUE) {

  if (!inherits(.df, "data.frame")) {
    stop("'.df' must be a data frame or tibble.")
  }

  group_expr   <- enquo(group_var)      # ???
  numeric_expr <- enquo(numeric_var)    # ???

  col_label <- paste0("mean_", as_label(numeric_expr))  # ???

  .df |>
    filter(!is.na(!!numeric_expr)) |>           # ???
    group_by(!!group_expr) |>                   # ???
    summarize(
      n             = n(),
      !!(col_label) := mean(!!numeric_expr, na.rm = na.rm)  # ???
    )
}

grouped_summary(penguins, species, body_mass_g)
grouped_summary(flights, carrier, dep_delay)
```
